Variational Monte Carlo on bosonic systems

### 2018-03-25 project_1/article/article.tex
Add typo in drift force
Add labels
Use small delta in time step
Replace tabs with spaces
	![](https://media1.giphy.com/media/l0IylSajlbPRFxH8Y/giphy.gif)

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -2\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the entire sum over all the random walkers
    disappears.

@@

        % TODO: Add the computation of this expression
        The final result for the energy is
        \begin{align}
            E(\alpha)
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
        \end{align}
        By minimizing this expression, i.e., finding the derivative of the
        energy with respect to $\alpha$ and equating this to zero, yields the
        expected minimum of variational energy to be
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}
        which in natural units reduces to $\alpha_0 = \half\omega$. The energy at
        this value of $\alpha$ (in natural units) is then
        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2}.
        \end{align}

@@

        Looking at the one-body part of the Hamiltonian we get
        \begin{align}
            H
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}\nabla_k^2
                + \half m\brak{
                    \omega^2(x_k^2 + y_k^2)
                    + \omega_z^2 z_k^2
                }
            }
            \\
            &=
            \frac{\hbar\omega}{2}
            \sum_{k = 1}^N
            \brac{
                -{\nabla_k'}^2
                +
                \brak{
                    {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                }
            },
        \end{align}
        where we have introduced the dimensionless frequency $\lambda =
        \omega_z/\omega$. The single particle functions also gets a scaling
        factor
        \begin{align}
            \phi_k
            &=
            \exp\brak{
                -\frac{\alpha\hbar}{m\omega}
                ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
            }
            \equiv
            \phi(\vf{r}_k').
        \end{align}
        The correlation wavefunction and the interaction potential remain the
        same, but using the scaled hard core sphere radius $a/a_{\text{ho}} =
        0.0043$. For brevity we will remove the ticks on the variables.

@@

            We use the Langevin equation to find the new position of the
            particle.
            \begin{align}
                \dpd[]{\vf{r}}{t}
                &=
                D\vf{F}(\vf{r}) + \vfg{\eta},
                \label{eq:langevin}
            \end{align}
            where $\vfg{\eta}$ is a uniformly distributed stochastic variable
            for each dimenion.  Solving Langevin's equation by Euler's method
            gives a recursive relation for the subsequent new positions of a
            particle.
            \begin{align}
                \vf{r}_{i + 1}
                &=
                \vf{r}_i + D\vf{F}(\vf{r}_i)\Delta t
                + \vfg{\xi}\sqrt{\Delta t},
            \end{align}
            for a given time step $\Delta t$
            \footnote{
                Bear in mind that \autoref{eq:langevin} is only valid as $\Delta
                t \to 0$, a property stemming from the use of Euler's method.
            }
            and a normally distributed stochastic variable $\vfg{\xi}$.

            Now we need to change the acceptance probability of the metropolis
            algorithm to something that takes the new sampling method into
            account.
            \begin{align}
                q(\vf{r}_{i + 1}, \vf{r}_i)
                &=
                \frac{
                    G(\vf{r}_{i + 1}, \vf{r}_{i}, \Delta t)
                    \left|\Psi_T(\vf{r}_{i + 1})\right|^2
                }
                {
                    G(\vf{r}_{i}, \vf{r}_{i + 1}, \Delta t)
                    \left|\Psi_T(\vf{r}_{i})\right|^2
                },
            \end{align}
            where $G$ is the Green's function of the Fokker-Planck equation
            given by
            \begin{align}
                G(\vf{r}_{i + 1}, \vf{r}_i, \Delta t)
                &=
                \exp\Biggl(
                    -\frac{\bigl[
                        \vf{r}_{i + 1}
                        - \vf{r}_{i}
                        - D\vf{F}(\vf{r}_i)\Delta t
                    \bigr]^2}{4D\Delta t}
                \Biggr)
                \nonumber \\
                &\qquad
                \times
                \frac{1}{(4\pi D\Delta t)^{dN/2}},
            \end{align}
            where $d$ is the dimensionality.

@@

		Scanning the entire space of variational parameters can be a somewhat despairing task. To aid us in the search we
		therefore implement the method of gradient descent\footnote{Also known as steepest descent, easily confused
		with the steepest descent method for integation.}. In any optimum we generally have that the derivative should equal
		zero. In our case, we want to find the minimum of the expected local energy with respect to some variational
		parameter $\alpha$, so that
		\begin{equation}
			\nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
		\end{equation}	
		The foundation for  the method of gradient descent lies in the fact that some analytic
		\footnote{We use the term analytic quite recklessly.} function $F(\alpha)$ decreases fastest in the direction
		of $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to an ensuing minimum by trailing along the
		pathway of the following recursive relation,
		\begin{equation}
			\alpha_{k+1} = \alpha_k - \gamma \nabla_{\alpha k} F(\alpha_k).
		\end{equation}
		If $\gamma$ is small enough we should have $f(\alpha_k) \geq F(\alpha_{k+1})$ for all $k$ and hopefully we will
		experience convergence to the desired minimum.

@@

>>>>

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the entire sum over all the random walkers
    disappears.

@@

        % TODO: Add the computation of this expression
        The final result for the energy is
        \begin{align}
            E(\alpha)
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
            \label{eq:exact_energy}
        \end{align}
        By minimizing this expression, i.e., finding the derivative of the
        energy with respect to $\alpha$ and equating this to zero, yields the
        expected minimum of variational energy to be
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}
        which in natural units reduces to $\alpha_0 = \half\omega$. The energy at
        this value of $\alpha$ (in natural units) is then
        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2}.
        \end{align}

@@

        Looking at the one-body part of the Hamiltonian we get
        \begin{align}
            H
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}\nabla_k^2
                + \half m\brak{
                    \omega^2(x_k^2 + y_k^2)
                    + \omega_z^2 z_k^2
                }
            }
            \\
            &=
            \frac{\hbar\omega}{2}
            \sum_{k = 1}^N
            \brac{
                -{\nabla_k'}^2
                +
                \brak{
                    {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                }
            },
            \label{eq:scaled_hamiltonian}
        \end{align}
        where we have introduced the dimensionless frequency $\lambda =
        \omega_z/\omega$. The single particle functions also gets a scaling
        factor
        \begin{align}
            \phi_k
            &=
            \exp\brak{
                -\frac{\alpha\hbar}{m\omega}
                ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
            }
            \equiv
            \phi(\vf{r}_k').
            \label{eq:scaled_spf}
        \end{align}
        The correlation wavefunction and the interaction potential remain the
        same, but using the scaled hard core sphere radius $a/a_{\text{ho}} =
        0.0043$. For brevity we will remove the ticks on the variables.

@@

            We use the Langevin equation to find the new position of the
            particle.
            \begin{align}
                \dpd[]{\vf{r}}{t}
                &=
                D\vf{F}(\vf{r}) + \vfg{\eta},
                \label{eq:langevin}
            \end{align}
            where $\vfg{\eta}$ is a uniformly distributed stochastic variable
            for each dimenion.  Solving Langevin's equation by Euler's method
            gives a recursive relation for the subsequent new positions of a
            particle.
            \begin{align}
                \vf{r}_{i + 1}
                &=
                \vf{r}_i + D\vf{F}(\vf{r}_i)\delta t
                + \vfg{\xi}\sqrt{\delta t},
            \end{align}
            for a given time step $\delta t$
            \footnote{
                Bear in mind that \autoref{eq:langevin} is only valid as $\Delta
                t \to 0$, a property stemming from the use of Euler's method.
            }
            and a normally distributed stochastic variable $\vfg{\xi}$.

            Now we need to change the acceptance probability of the metropolis
            algorithm to something that takes the new sampling method into
            account.
            \begin{align}
                q(\vf{r}_{i + 1}, \vf{r}_i)
                &=
                \frac{
                    G(\vf{r}_{i + 1}, \vf{r}_{i}, \delta t)
                    \left|\Psi_T(\vf{r}_{i + 1})\right|^2
                }
                {
                    G(\vf{r}_{i}, \vf{r}_{i + 1}, \delta t)
                    \left|\Psi_T(\vf{r}_{i})\right|^2
                },
            \end{align}
            where $G$ is the Green's function of the Fokker-Planck equation
            given by
            \begin{align}
                G(\vf{r}_{i + 1}, \vf{r}_i, \delta t)
                &=
                \exp\Biggl(
                    -\frac{\bigl[
                        \vf{r}_{i + 1}
                        - \vf{r}_{i}
                        - D\vf{F}(\vf{r}_i)\delta t
                    \bigr]^2}{4D\delta t}
                \Biggr)
                \nonumber \\
                &\qquad
                \times
                \frac{1}{(4\pi D\delta t)^{dN/2}},
            \end{align}
            where $d$ is the dimensionality.

@@

        Scanning the entire space of variational parameters can be a somewhat
        despairing task. To aid us in the search we therefore implement the
        method of gradient descent\footnote{Also known as steepest descent,
        easily confused with the steepest descent method for integation.}. In
        any optimum we generally have that the derivative should equal zero. In
        our case, we want to find the minimum of the expected local energy with
        respect to some variational parameter $\alpha$, so that
        \begin{equation}
            \nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
        \end{equation}
        The foundation for  the method of gradient descent lies in the fact that
        some analytic \footnote{We use the term analytic quite recklessly.}
        function $F(\alpha)$ decreases fastest in the direction of
        $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to
        an ensuing minimum by trailing along the pathway of the following
        recursive relation,
        \begin{equation}
            \alpha_{k+1} = \alpha_k - \gamma \nabla_{\alpha} F(\alpha_k).
        \end{equation}
        If $\gamma$ is small enough we should have $f(\alpha_k) \geq
        F(\alpha_{k+1})$ for all $k$ and hopefully we will experience
        convergence to the desired minimum.

@@

	\subsubsection{Test of Gradient descent}

	Before moving on to the interacting, more interesting systems, it would be beneficial to test the gradient descent implementation
	now that we are certain that it moves towards the optimum variational parameter $\alpha = \frac{1}{2}$.
